### 2018-05-30 latex/sc18/abstract.tex
put changes in red

The CP tensor decomposition is a low-rank approximation of a tensor. 
We present a distributed-memory parallel algorithm and implementation of an alternating optimization method for computing a CP decomposition of dense tensor data that can enforce nonnegativity of the computed low-rank factors.
The principal task is to parallelize the matricized-tensor times Khatri-Rao product (MTTKRP) bottleneck subcomputation.
The algorithm is computation efficient, using dimension trees to avoid redundant computation across MTTKRPs within the alternating method.
Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to 100s of nodes (up to 8192 cores) and is faster and more general than the currently available parallel software.

>>>>

The CP tensor decomposition is a low-rank approximation of a tensor. 
We present a distributed-memory parallel algorithm and implementation of an alternating optimization method for computing a CP decomposition of dense tensor data that can enforce nonnegativity of the computed low-rank factors.
The principal task is to parallelize the matricized-tensor times Khatri-Rao product (MTTKRP) bottleneck subcomputation.
The algorithm is computation efficient, using dimension trees to avoid redundant computation across MTTKRPs within the alternating method.
Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to 100s of nodes (up to \red{4096} cores) and is faster and more general than the currently available parallel software.

### 2018-05-30 latex/sc18/algorithm.tex
put changes in red

To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
Then 
\begin{equation*}
\MnE{M}{1}{ir} = \sum_{i,j} \TE{X}{ijk} \ME{V}{jr} \ME{W}{kr} 
= \sum_{j} \ME{V}{jr} \sum_k \TE{X}{ijk} \ME{W}{kr} 
= \sum_{j} \ME{V}{jr} \TE{T}{ijr},
\end{equation*}
where $\T{T}$ is an $I\times J\times R$ tensor that is the result of a partial MTTKRP between tensor $\T{X}$ and the single factor matrix $W$.
Likewise,
\begin{equation*}
\MnE{M}{2}{jr} = \sum_{i,k} \TE{X}{ijk} \ME{U}{ir} \ME{W}{kr} 
= \sum_{i} \ME{U}{ir} \sum_k \TE{X}{ijk} \ME{W}{kr} 
= \sum_{i} \ME{U}{ir} \TE{T}{ijr},
\end{equation*}
and we see that the temporary tensor $\T{T}$ can be re-used.
From these expressions, we can also see that computing $\T{T}$ (a partial MTTKRP) corresponds to a matrix-matrix multiplication, and computing each of $\Mn{M}{1}$ and $\Mn{M}{2}$ from $\T{T}$ (a multi-TTV) corresponds to $R$ independent matrix-vector multiplications.
In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.

@@

\Cref{fig:DT} illustrates a dimension tree for the case $N=5$.
Various shapes of binary trees are possible \cite{PTC13a,Kaya17}.
For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
In this way, each partial MTTKRP can be performed via BLAS's GEMM interface without reordering tensor entries in memory.
All other edges in a tree correspond to multi-TTVs and are typically much cheaper.
By organizing the memory layout of temporary quantities, the multi-TTV operations can be performed via a sequence of calls using BLAS's GEMV interface.
By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.

\begin{figure}
\subfloat[Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call. \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
\subfloat[Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.  Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$. \label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
\caption{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.  In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}
\label{fig:DTcartoon}
\end{figure}

\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
\Cref{fig:DT-PM} shows a partial MTTKRP between the input tensor $\T{X}$ and the Khatri-Rao product of the factor matrices in modes 1 and 2, which produces a temporary tensor $\T{T}$ corresponding to the $\{3,4,5\}$ node in the dimension tree.
The key to efficiency in this computation is that the matricization of $\T{X}$ that assigns modes 1 through 2 to rows and modes 3 through 5 to columns is already column-major in memory.
Thus, we can use the GEMM interface and compute the temporary tensor $\T{T}$ without reordering any tensor entries.
\Cref{fig:DT-mTTV} depicts a multi-TTV that computes the results $\Mn{M}{3}$ from $\T{T}$ and the factor matrices in modes 4 and 5.
Here, the tensor $\T{T}$ is matricized with respect to only its first mode (of dimension $I_3$), but this matricization is also column-major in memory.
We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.

No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
This is the same operation count as a full MTTKRP.
The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.

The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.

@@

The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
The main idea is that at the start of the $n$th inner iteration (\cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
After the local MTTKRP is computed (\cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\cref{fig:inner_d}).
The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.

@@

\begin{table*}
\centering
\begin{tabular}{|c|cccc|}
\hline
 & \textbf{Computation} & \textbf{Memory - Local MTTKRP} & \textbf{Communication} & \textbf{Memory - Par.~Algorithm}  \\
\hline
Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
%\hline
Par-NNCP w/ gen.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$  \\
%\hline
Par-NNCP w/o dim.~tree opt. & $\mathbf{\frac{NIR}{P}}$ & $\displaystyle \mathbf{\max_n \frac{RI/I_n}{P/P_n}}$  &  $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ \\
\hline
\end{tabular}
\smallskip
\caption{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).  We ignore constants but omit big-Oh notation for clarity.  The first line corresponds to \cref{alg:Par-NNCP-short} with an optimal choice of processor grid and applying the dimension tree optimization locally.  The second line corresponds to the changes in communication and memory for a general processor grid.  The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}
\label{tab:costs}
\end{table*}

We will analyze the cost of a single outer iteration.
While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.

\paragraph{\emph{Computation}}
The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$, which is dominated by the cost of the MTTKRP.
When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.

>>>>

\red{To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
Then 
\begin{equation*}
\MnE{M}{1}{ir} = \sum_{i,j} \TE{X}{ijk} \ME{V}{jr} \ME{W}{kr} 
= \sum_{j} \ME{V}{jr} \sum_k \TE{X}{ijk} \ME{W}{kr} 
= \sum_{j} \ME{V}{jr} \TE{T}{ijr},
\end{equation*}
where $\T{T}$ is an $I\times J\times R$ tensor that is the result of a partial MTTKRP between tensor $\T{X}$ and the single factor matrix $W$.
Likewise,
\begin{equation*}
\MnE{M}{2}{jr} = \sum_{i,k} \TE{X}{ijk} \ME{U}{ir} \ME{W}{kr} 
= \sum_{i} \ME{U}{ir} \sum_k \TE{X}{ijk} \ME{W}{kr} 
= \sum_{i} \ME{U}{ir} \TE{T}{ijr},
\end{equation*}
and we see that the temporary tensor $\T{T}$ can be re-used.
From these expressions, we can also see that computing $\T{T}$ (a partial MTTKRP) corresponds to a matrix-matrix multiplication, and computing each of $\Mn{M}{1}$ and $\Mn{M}{2}$ from $\T{T}$ (a multi-TTV) corresponds to $R$ independent matrix-vector multiplications.
In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.}

@@

\Cref{fig:DT} illustrates a dimension tree for the case $N=5$.
Various shapes of binary trees are possible \cite{PTC13a,Kaya17}.
For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
\red{We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
In this way, each partial MTTKRP can be performed via BLAS's GEMM interface without reordering tensor entries in memory.
All other edges in a tree correspond to multi-TTVs and are typically much cheaper.
By organizing the memory layout of temporary quantities, the multi-TTV operations can be performed via a sequence of calls using BLAS's GEMV interface.
By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.}

\begin{figure}
\subfloat[\red{Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call.} \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
\subfloat[\red{Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.  Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$.} \label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
\caption{\red{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.  In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}}
\label{fig:DTcartoon}
\end{figure}

\red{\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
\Cref{fig:DT-PM} shows a partial MTTKRP between the input tensor $\T{X}$ and the Khatri-Rao product of the factor matrices in modes 1 and 2, which produces a temporary tensor $\T{T}$ corresponding to the $\{3,4,5\}$ node in the dimension tree.
The key to efficiency in this computation is that the matricization of $\T{X}$ that assigns modes 1 through 2 to rows and modes 3 through 5 to columns is already column-major in memory.
Thus, we can use the GEMM interface and compute the temporary tensor $\T{T}$ without reordering any tensor entries.
\Cref{fig:DT-mTTV} depicts a multi-TTV that computes the results $\Mn{M}{3}$ from $\T{T}$ and the factor matrices in modes 4 and 5.
Here, the tensor $\T{T}$ is matricized with respect to only its first mode (of dimension $I_3$), but this matricization is also column-major in memory.
We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.}

\red{No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
This is the same operation count as a full MTTKRP.
The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.}

\red{The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.}

@@

The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
The main idea is that at the start of the $n$th inner iteration (\Cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
After the local MTTKRP is computed (\Cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \Cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\Cref{fig:inner_d}).
The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\Cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.

@@

\begin{table*}
\centering
\begin{tabular}{|c|cccc|}
\hline
 & \textbf{Computation} & \textbf{Memory - Local MTTKRP} & \textbf{Communication} & \textbf{Memory - Par.~Algorithm}  \\
\hline
Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
%\hline
Par-NNCP w/ gen.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$  \\
%\hline
Par-NNCP w/o dim.~tree opt. & $\mathbf{\frac{NIR}{P}}$ & $\displaystyle \mathbf{\max_n \frac{RI/I_n}{P/P_n}}$  &  $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ \\
\hline
\end{tabular}
\smallskip
\caption{\red{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).  We ignore constants but omit big-Oh notation for clarity.  The first line corresponds to \cref{alg:Par-NNCP-short} with an optimal choice of processor grid and applying the dimension tree optimization locally.  The second line corresponds to the changes in communication and memory for a general processor grid.  The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}}
\label{tab:costs}
\end{table*}

We will analyze the cost of a single outer iteration.
While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
\red{We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.}

\paragraph{\emph{Computation}}
The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
The sum of these three costs across all inner iterations is \red{$O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$}, which is dominated by the cost of the MTTKRP.
When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.

### 2018-05-30 latex/sc18/conclusion.tex
put changes in red

In particular, the performance results demonstrate that computing NNCP for dense tensors involves heavy computation relative to the sizes of the computed factor matrices.
By avoiding the communication of tensor entries and communicating only the factor matrices, the parallel algorithm is nearly always compute bound.
This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
In the case of NNCP, the ratio of local computation to communication is $O((I/P)^{1-1/N}/N)$, which is approximately the size of the local data raised to the power $1-1/N$.
This exponent is larger than $1/2$ and grows with $N$, and therefore it predicts the NNCP computation should be more computation bound than matrix multiplication.
Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.

We can also conclude from the performance results that the dimension tree optimization is the key to performance improvement over the state-of-the-art approaches.
For 3D tensors, we observe a benefit larger than the theoretical 50\% reduction in computation, and for larger numbers of modes, the improvement is only magnified.
Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
Furthermore, we see that tuning the processor grid had much less effect on overall performance.
Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication.

>>>>

In particular, the performance results demonstrate that computing NNCP for dense tensors involves heavy computation relative to the sizes of the computed factor matrices.
By avoiding the communication of tensor entries and communicating only the factor matrices, the parallel algorithm is nearly always compute bound.
This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
\red{For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
In the case of NNCP, the ratio of local computation to communication is $O((I/P)^{1-1/N}/N)$, which is approximately the size of the local data raised to the power $1-1/N$.
This exponent is larger than $1/2$ and grows with $N$, and therefore it predicts the NNCP computation should be more computation bound than matrix multiplication.
Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.}

We can also conclude from the performance results that the dimension tree optimization is the key to performance improvement over the state-of-the-art approaches.
For 3D tensors, we observe a benefit larger than the theoretical 50\% reduction in computation, and for larger numbers of modes, the improvement is only magnified.
Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
\red{Furthermore, we see that tuning the processor grid had much less effect on overall performance.
Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication. }

### 2018-05-30 latex/sc18/experiments.tex
put changes in red

The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
There is little difference in performance between our implementation without dimension trees and NbAO-NTF.

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_3D_syn.dat}
\renewcommand{\numiterations}{1}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
\label{fig:strongsynthetic3D}
\end{figure}

@@

\subsection{Varying Processor Grid}

In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.

\begin{figure}
\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}
\begin{axis}[	
	ybar stacked,
	bar width=8pt,
	width=\columnwidth,
	height =.75\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Processor Grid},
	y label style={yshift=-.5cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
	%xtick=data,
	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
\label{fig:commsweep10}
\end{figure}

>>>>

The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
\red{Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.}
Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
There is little difference in performance between our implementation without dimension trees and NbAO-NTF.

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_3D_syn.dat}
\renewcommand{\numiterations}{1}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for \red{$k\in\{0,1,2,3,4\}$}.  The rank is fixed at 32.}
\label{fig:strongsynthetic3D}
\end{figure}

@@

\subsection{\red{Varying Processor Grid}}

\red{In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.}

\begin{figure}
\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}
\begin{axis}[	
	ybar stacked,
	bar width=8pt,
	width=\columnwidth,
	height =.75\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Processor Grid},
	y label style={yshift=-.5cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
	%xtick=data,
	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
\caption{\red{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}}
\label{fig:commsweep10}
\end{figure}
